---
title: "Assignment 1: Task 1 - CFA"
author: "Group 11"
date: ""
output: pdf_document
---
\section{0. Load Libraries}

```{r, message=FALSE, warning=FALSE}
library(lavaan)
library(semPlot)
```


\section{1. Problem Statement}
- model hypothesized
- nr of factors, which variables should load on which factor 

\section{2. Descriptive Statistics}
```{r}
cov <-read.delim("../data/basketball.txt", header = TRUE, sep="",dec = ".", skipNul = FALSE)
cov <- cov[-1]
cov <- cov[-c(1,2,3),]
rownames(cov) = c("X1","X2","X3","X4","X5","X6","X7","X8")
colnames(cov) = c("X1","X2","X3","X4","X5","X6","X7","X8")
cov <- as.matrix(cov)
```


\section{3. Assumptions}
• check whether the model is identified
(otherwise there are no unique parameter estimates)
a necessary condition for the model to be identified:
the nr of parameters to be estimated ≤ the nr of inputs
• nr of inputs = nr of unique elements in the covariance matrix
• for example: if you have 3 observed variables (indicators)
you can compute 3 variances and 3 covariances
⇒ the number of inputs = 6
(this is independent of the number of observations!)
• in general: for p observed variables
covariances
you can compute p variances and p(p−1)/2

⇒ number of inputs = p(p+1)/2

36 inputs
p = 8 -> 8 loadings, 8 error variances, 1 covariance -> 17
17<36 model is overidentified -> good 


\section{4. Method}


• if the model is identified ⇒ estimate the parameters
• how well does the model fit the data?
– look at goodness of fit indices
– if the model fits the data ⇒ evaluate the estimates
– if the model does not fit the data ⇒ check whether the model
can be improved (model respecification).
```{r}
model <- '
# latent variables
F1 =~ X1 + X2 + X3 + X4 
F2 =~ X5 + X6 + X7 + X8 '
fit <- lavaan::cfa(model,sample.nobs = 200,sample.cov=cov)

summary(fit, fit.measures=TRUE)


```
\section{5. Interpretation}

GFI = model fit statistic = 71.186 should be > 0.95
CFI = 0.903 should be >0.95
P value for chi^2 = 0.000 -> should be > 0.05
AIC 1413 -> as small as possible

no, model does not fit the data

different model, now 3 factors, they correlate -> 3 variances -> 19  parameters instead of 17 

```{r}
model <- '
# latent variables
F1 =~ X1 + X2 
F2 =~ X3 + X4
F3 =~ X5 + X6 + X7 + X8 '

fit2 <- lavaan::cfa(model,sample.nobs = 200,sample.cov=cov)

summary(fit2, fit.measures=TRUE)
loadings2 <- fit2@Model
fit2@ParTable

```

GFI 34. -> worse?
P = 0.008 < 0.05
CFI good
SRMR (Standardized residual) should not be larger than 1.96
 R^2? should be > 0.5 
